Trajectory prediction systems are evaluated primarily by geometric coverage, yet deployment depends on the single trajectory ultimately selected. When multiple plausible futures exist, choosing by confidence alone can preserve geometric plausibility while still selecting a safety- or law-violating mode. This paper studies that selection step directly.

\textsc{RECTOR} (\textbf{R}ule-\textbf{E}nforced \textbf{C}onstrained \textbf{T}rajectory \textbf{OR}chestrator) is a rule-aware selection layer that ranks candidate ego trajectories under the priority order Safety~$\succ$~Legal~$\succ$~Road~$\succ$~Comfort. It combines scene-conditioned rule activation, differentiable rule proxies, and a deterministic lexicographic selector that operates entirely after generation. On WOMD \texttt{validation\_interactive}, rule-aware selection raises selected-mode compliance from 76.14\% to 84.97\% (23.86\%\,$\to$\,15.03\% total violations), with the gain concentrated in Safety-tier filtering. Since the benchmark is already largely compliant, the practical effect is avoiding the minority of high-risk modes in scenes where the candidate set contains both compliant and non-compliant options. Under the reported evaluation, weighted-sum and lexicographic selectors produce the same binary compliance outcomes; the lexicographic selector's value lies in its formal priority guarantee, parameter-free behavior, and auditability. The paper reports Protocol~A as the operating-stack selector comparison and Protocol~B with conservative Safety/Legal activation as the stricter reference for absolute compliance.

%-------------------------------------------------------------------------------
\subsection{Motivation: The Compliance Gap}
\label{subsec:motivation}
%-------------------------------------------------------------------------------

Standard best-of-$K$ metrics such as minADE measure candidate-set coverage, not the acceptability of the executed mode. In interactive scenes, two candidates can be geometrically similar while differing sharply in collision risk, signal compliance, or right-of-way behavior. This gap is amplified in the heavy-tailed error regime of WOMD \texttt{validation\_interactive}, where the median minADE is 0.46\,m but the 95th and 99th percentiles rise to 2.16\,m and 3.62\,m respectively.

%-------------------------------------------------------------------------------
\subsection{Key Technical Observations}
\label{subsec:observations}
%-------------------------------------------------------------------------------

RECTOR is motivated by four observations. First, traffic rules are hierarchical: safety dominates legality, which dominates road-level and comfort criteria. Second, rule applicability is scene-dependent, so irrelevant rules should not influence ranking. Third, many high-value rules admit smooth approximations that are stable enough for batched ranking. Fourth, finite candidate sets may contain no fully compliant option, so the selector must degrade gracefully to the least-violating candidate rather than reverting to confidence alone.

%-------------------------------------------------------------------------------
\subsection{Approach Overview}
\label{subsec:approach}
%-------------------------------------------------------------------------------

In the reference stack used here, a Transformer--CVAE generator inspired by M2I produces $K{=}6$ candidate ego trajectories over a 5\,s horizon ($T{=}50$ timesteps at 10\,Hz). RECTOR then evaluates each candidate against a catalog of $R{=}28$ rules organized into four tiers:
\begin{itemize}[leftmargin=1.5em] 

\item \textbf{Safety (Tier~0):} collision avoidance and minimum distance
constraints (5~rules), 

\item \textbf{Legal (Tier~1):} traffic-signal and stop-sign compliance, speed limits, and right-of-way rules (7~rules),\footnote{Tier labels are used as research abstractions within the WOMD-based rule catalog; deployment and jurisdictional qualifications are discussed in Sections~\ref{Sec:Known_Gaps_and_Next_Steps_Toward_Deployment} and~\ref{Sec:Conclusion}.}


\item \textbf{Road (Tier~2):} lane keeping and road boundary adherence
(2~rules), \item \textbf{Comfort (Tier~3):} acceleration, jerk, and steering bounds
and smooth motion preferences (14~rules).

\end{itemize}

Candidates are ranked lexicographically: Safety first, then Legal, then Road, then Comfort. The selector is deterministic and auditable, and it leaves the upstream generator unchanged.

%-------------------------------------------------------------------------------
\subsection{Contributions}
\label{subsec:contributions}
%-------------------------------------------------------------------------------

The paper makes five contributions:

\begin{enumerate}[label=\textbf{C\arabic*},leftmargin=2.5em]

\item \textbf{Rule-aware selection improves Safety filtering on a fixed candidate set.} On 12{,}800 WOMD scenarios, reranking raises selected-mode compliance from 76.14\% to 84.97\% and reduces Safety$+$Legal violations from 21.87\% to 13.15\%.

\item \textbf{Lexicographic selection provides a structural guarantee.} Across the tested weight grids, weighted-sum matches lexicographic binary compliance; the lexicographic selector is valuable because it is parameter-free, auditable, and cannot trade higher-priority violations for lower-priority gains.

\item \textbf{Conservative always-on activation outperforms the learned applicability head for Safety and Legal tiers.} Cross-evaluation on 43{,}219 scenarios shows that always-on activation improves both compliance and accuracy, making the learned head a negative result for high-priority tiers.

\item \textbf{Formal properties and empirical stress tests.} We prove order invariance, infeasibility fallback, and scalarization correctness for the selector, and verify semantic invariants, numerical stability, and proxy--auditor alignment for the evaluation stack.

\item \textbf{Practical single-GPU implementation.} Mean inference latency is 7.3\,ms per sample; a 55-scenario Waymax log-replay bridge confirms end-to-end execution.

\end{enumerate}

%-------------------------------------------------------------------------------
\subsection{System Architecture}
\label{subsec:architecture}
%-------------------------------------------------------------------------------

RECTOR augments a standard multi-modal generator with two rule-aware components (Fig.~\ref{fig:rector_architecture}): an applicability mechanism that activates scene-relevant rules, and a tiered scorer that selects the candidate with the smallest prioritized violation vector. The selector itself contains no learned parameters and can be attached to any upstream generator.

%-------------------------------------------------------------------------------
\subsection{Interpretability and Explainability}
\label{subsec:interpretability}
%-------------------------------------------------------------------------------

The selection step is auditable by construction. For each decision, RECTOR records the active rules, tier scores, and the first tier that separated the selected candidate from its alternatives. This trace is deterministic and reproducible, even though the learned applicability head, when used, remains an upstream component.

%-------------------------------------------------------------------------------
\subsection{Operating Setup}
\label{subsec:scope}
%-------------------------------------------------------------------------------

\textsc{RECTOR} performs open-loop ego trajectory selection over a 5-second horizon ($T{=}50$ timesteps at 10\,Hz, $K{=}6$ modes). The paper, therefore, addresses tactical compliance within that horizon; route-level and longer-horizon planning remain outside the scope of the present paper. The reported gains are selection-time gains on the same candidate set, not claims that the benchmark trajectories are broadly unsafe or that the generator itself has been retrained into a new behavior regime.

%===============================================================================
%  FIGURES
%===============================================================================

% — Figure 1: RECTOR System Overview (full-width TikZ) —
\begin{figure*}[t]
\centering
\begin{tikzpicture}[
% Node styles
input/.style={
rectangle,
draw=gray!70,
fill=gray!10,
minimum width=2.2cm,
minimum height=0.7cm,
font=\footnotesize,
align=center
},
encoder/.style={
rectangle,
rounded corners=3pt,
draw=blue!70!black,
fill=blue!15,
minimum width=2.4cm,
minimum height=1.8cm,
font=\small\bfseries,
align=center
},
decoder/.style={
rectangle,
rounded corners=3pt,
draw=green!60!black,
fill=green!15,
minimum width=2.4cm,
minimum height=1.8cm,
font=\small\bfseries,
align=center
},
rulehead/.style={
rectangle,
rounded corners=3pt,
draw=orange!80!black,
fill=orange!15,
minimum width=2.2cm,
minimum height=1.4cm,
font=\small\bfseries,
align=center
},
proxy/.style={
rectangle,
rounded corners=3pt,
draw=red!70!black,
fill=red!10,
minimum width=2.2cm,
minimum height=1.4cm,
font=\small\bfseries,
align=center
},
scorer/.style={
rectangle,
rounded corners=3pt,
draw=purple!70!black,
fill=purple!15,
minimum width=2.2cm,
minimum height=1.4cm,
font=\small\bfseries,
align=center
},
output/.style={
rectangle,
draw=gray!70,
fill=yellow!10,
minimum width=2.2cm,
minimum height=0.7cm,
font=\footnotesize,
align=center
},
data/.style={
rectangle,
draw=gray!50,
fill=white,
minimum width=1.8cm,
minimum height=0.6cm,
font=\scriptsize,
align=center
},
arrow/.style={
->,
>=Stealth,
thick,
gray!70
},
dataarrow/.style={
->,
>=Stealth,
semithick,
gray!50
},
label/.style={
font=\scriptsize\itshape,
text=gray!70
}
]

% ============================================
% ROW 1: Main prediction pipeline
% ============================================

% Inputs
\node[input] (ego) at (0, 0) {Ego History\\$[\text{B}, 11, 4]$};
\node[input] (agents) at (0, -1) {Agent States\\$[\text{B}, 32, 11, 4]$};
\node[input] (lanes) at (0, -2) {Lane Centers\\$[\text{B}, 64, 20, 2]$};

% Input group box
\begin{scope}[on background layer]
\node[draw=gray!40, dashed, rounded corners=5pt,
fit=(ego)(agents)(lanes),
inner sep=6pt,
label={[font=\footnotesize\bfseries, text=gray!60]above:Inputs}] (inputbox) {};
\end{scope}

% M2I Encoder
\node[encoder, right=0.5cm of agents] (m2i) {M2I-style\\Scene Encoder\\[2pt]\footnotesize(PointNet + Transformer)};
\node[label, below=0.5cm of m2i] {\textbf{jointly trained}};

% Scene embedding (intermediate)
\node[data, below=0.5cm of m2i] (scene_emb) {Scene Emb.\\$[\text{B}, 256]$};

% CVAE Decoder
\node[decoder, right=0.5cm of scene_emb] (cvae) {CVAE\\Decoder};

% Trajectory outputs
\node[output, right=0.6cm of cvae, yshift=0.4cm] (traj) {Ego Trajectories\\$[\text{B}, 6, 50, 4]$};
\node[output, right=0.6cm of cvae, yshift=-0.4cm] (conf) {Confidences\\$[\text{B}, 6]$};

% ============================================
% ROW 2: Rule evaluation pipeline
% ============================================

% Rule Applicability Head
\node[rulehead, below=1.5cm of scene_emb] (apphead) {Rule\\Applicability\\Head};

% Applicability output
\node[data, right=0.8cm of apphead] (app_out) {Applicability\\$\mathbf{a} \in [0,1]^{28}$};

% Smooth Rule Proxies
\node[proxy, right=1.0cm of app_out, yshift=0cm] (proxies) {Smooth\\Rule Proxies};

% Violations
\node[data, below=1.0cm of proxies] (viol) {Violations\\$[\text{B}, 6, 24]$\\{\footnotesize$^*$}};

% Tiered Scorer
\node[scorer, right=0.8cm of viol] (scorer) {Tiered\\Lexicographic\\Scorer};

% Final output
\node[output, right=1.0cm of scorer, yshift=0.3cm] (selected) {Selected Ego\\Trajectory};
\node[output, right=1.0cm of scorer, yshift=-0.5cm] (explain) {Rule-based\\Explanation};

% Inputs to encoder
\draw[arrow] (ego.east) --++(0.3,0) |- ([yshift=0.5cm]m2i.west);
\draw[arrow] (agents.east) --(m2i.west);
\draw[arrow] (lanes.east) --++(0.3,0) |- ([yshift=-0.5cm]m2i.west);

% Encoder to scene embedding
\draw[arrow] (m2i.south) --(scene_emb.north);

% Scene embedding to decoder
\draw[arrow] (scene_emb.east) --(cvae.west);

% Decoder to outputs
\draw[arrow] (cvae.east) --++(0.3,0) |- (traj.west);
\draw[arrow] (cvae.east) --++(0.3,0) |- (conf.west);

% ============================================
% ARROWS - Rule pipeline
% ============================================

% Scene embedding to applicability head (branch down)
\draw[arrow] (scene_emb.south) --(apphead.north);

% Applicability head to output
\draw[arrow] (apphead.east) --(app_out.west);

% Trajectories down to proxies
\draw[arrow] (conf.south) --++(0, -0.5) -| ([xshift=-0.3cm]proxies.north);

% Applicability to proxies
\draw[arrow] (app_out.east) --([yshift=0cm]proxies.west);

% Context (lanes) to proxies - dashed for context
\draw[dataarrow, dashed] (lanes.south) --++(0, -2.0) -| ([xshift=-0.5cm]proxies.north);
\node[label, below=1.5cm of lanes, xshift=1.5cm] {\textbf{map context}};

% Proxies to violations
\draw[arrow] (proxies.south) --(viol.north);

% Violations to scorer
\draw[arrow] (viol.east) --(scorer.west);

% Confidences to scorer
\draw[arrow] (conf.east) --++(0.0, 0.0) -| (scorer.north);

% Scorer to outputs
\draw[arrow] (scorer.east) --++(0.3,0) |- (selected.west);
\draw[arrow] (scorer.east) --++(0.3,0) |- (explain.west);

\node[font=\scriptsize, text=purple!70!black, align=left, anchor=north west]
at ([xshift=-0.7cm, yshift=-0.3cm]scorer.south west) {
Safety $\succ$ Legal $\succ$ Road $\succ$ Comfort
};
\end{tikzpicture}
\caption{\textbf{RECTOR system overview.}  Scene inputs (ego history, agent states, lane geometry) flow through an M2I-style encoder (trained jointly with the decoder) to produce a dense scene embedding
$\mathbf{e}_{\text{scene}}\in\mathbb{R}^{256}$.  The Transformer–CVAE decoder generates six diverse ego-trajectory hypotheses, each with an associated confidence score.  In parallel, the \emph{rule applicability head} (the core learned component; 3.33\,M of 8.83\,M total parameters) predicts which of 28~traffic rules are relevant to the current scenario.  Smooth proxies compute violation severities for each trajectory–rule pair ($^*$24/28~rules have differentiable proxies; the remaining 4 are evaluated post-selection via the full-catalog auditor and contribute zero to proxy-based scores), and the tiered lexicographic scorer selects the ego trajectory with minimal violations respecting strict tier priority
(Safety~$\succ$~Legal~$\succ$~Road~$\succ$~Comfort).  The selected trajectory includes a rule-based explanation of the selection decision.}
\label{fig:rector_architecture}
\end{figure*}
