We introduced ¥textsc{RECTOR}, a rule-aware trajectory selector that operates on a fixed multi-modal candidate set and enforces the priority order Safety‾$¥succ$‾Legal‾$¥succ$‾Road‾$¥succ$‾Comfort at selection time. On 12{,}800 WOMD ¥texttt{validation¥_interactive} scenarios, rule-aware reranking raises reported Protocol‾A selected-mode compliance from 76.14¥% to 84.97¥%, driven by Safety-tier filtering. The lexicographic selector adds a parameter-free decision rule with explicit priority guarantees, and the implementation remains practical at the single-GPU scale, with a mean inference latency of 7.3¥,ms.

Two secondary findings deserve emphasis. First, the learned applicability head yields a negative result for the Safety and Legal tiers: conservative, always-on activation performs better in cross-evaluation and should be the default policy. Second, the proxy stack is good enough for selection-time ranking but not a substitute for audit: pairwise candidate-ranking concordance is high, while some per-rule severity correlations remain weak.

The present limits are clear. Results are open-loop, tied to a single generator, and evaluated with U.S.-centric rules over a 5¥,s horizon. For absolute compliance, the conservative reference is Protocol‾B together with always-on Safety/Legal activation: oracle applicability exposes materially higher violation rates than the learned head, four rules remain audit-only, and the current closed-loop study validates the bridge rather than a reactive safety stack. Within those limits, RECTOR shows that explicit, priority-ordered selection is a practical and auditable mechanism for translating multimodal diversity into improved compliance.
